---
title: "Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Follow this command to install the package:

```{r eval=FALSE}

if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
library(devtools)
devtools::install_github("https://github.com/elenasasso/GenesClasses")
```

```{r setup, warning=FALSE, message=FALSE}
library(GenesClasses)
```

# Description

This vignette helps to understand how to use the `GenesClasses` package.

The package contains a virtual general class called "Gene," from which all other specific classes derive. This class includes general attributes that are shared among all genes.

The specific gene classes are divided into three main categories: coding genes, Housekeeping RNA genes, and non-coding RNA genes.

The second and third categories are further subdivided into more specific classes:

-   Housekeeping RNA genes are divided into tRNA genes and rRNA genes.

-   Non-coding RNA genes are divided into long non-coding RNA genes and short non-coding RNA genes, with the latter further subdivided into miRNA, snoRNA, and snRNA.

For each gene, it is possible to compute the length of the gene in base pairs and the length of the product it generates.

Below are some simple examples for all the classes provided in the package and for all the functions.

# Classes:

## Gene

The gene class is a generic virtual class for which is not possible to create object, but it contains all the attribute shared among all genes.

**NB**: Remember not to try to create object of this class: the example below is wrong!

```{r eval= FALSE}
g1 <- Gene(id = 1)
```

## Coding gene

This class contains protein-coding genes, inheriting all attributes from the Gene class, and includes additional attributes specific to protein-coding genes (e.g., the coordinates of the CDS).

```{r}
brca1 <- CodingGene(id = 207233, Hugo_symbol = 'BRCA1', chrom = 'chr17', CDS_end = 41197695, CDS_start = 41258496, strand = '-')
```

```{r}
brca2 <- CodingGene(id = 563829, Hugo_symbol = 'BRCA2', chrom = 'chr13', CDS_start = 32890598, CDS_end = 32972907, strand = '+')
```

## Housekeeping RNA genes

It's actually possible to create object of this class if you just know that the gene is an housekeeping RNA but you don't know if it's a tRNA or an rRNA, but it's more interesting to see example of the specific case.

```{r}
HkGene <- Housekeeping_RNA_Gene(id = 14567, category = "Housekeeping")
```

### tRNA

**NB**: Be carefull to write the correct amminoacids: the first example below is wrong because the "e" at the end of the name is missing.

```{r eval=FALSE}
tRNA1 <- tRNA_Gene(id = 4672829, start = 3562, end = 4567,number_exon = 8,  specific_aa = "Alanin")
```

```{r}
tRNA1 <- tRNA_Gene(id = 4672829, start = 3562, end = 4567,number_exon = 8,  specific_aa = "Alanine")
```

```{r}
tRNA2 <- tRNA_Gene(id = 46271, number_exon = 4,  specific_aa = "Glutamine")
```

```{r}
tRNA3 <- tRNA_Gene(id = 10087, start = 1578, end = 2766, number_exon = 10,  specific_aa = "Tyrosine")
```

### rRNA

**NB**: Only eukaryotic rRNAs are included! The first example below is incorrect because 16s is an rRNA for prokaryotes.

```{r eval=FALSE}
rRNA1 <- rRNA_Gene(id = 627291, rRNA_ID = 356282,  specific_RNA = "16s")
```

```{r}
rRNA1 <- rRNA_Gene(id = 627291, rRNA_ID = 356282,  specific_RNA = "18s")
```

```{r}
rRNA2 <- rRNA_Gene(id = 98578, start = 282, end = 876, specific_RNA = "5.8s")
```

```{r}
rRNA3 <- rRNA_Gene(id = 268191,  TSS = 886, TTS = 6567 ,specific_RNA = "5s")
```

```{r}
rRNA4 <- rRNA_Gene(id = 2681, length_rRNA = 87, rRNA_ID = 350,  specific_RNA = "28s")
```

## Non-coding genes

It's actually possible to create object of this class if you just know that the gene is a regolative non-coding RNA but you don't know if it's a lncRNA or sncRNA, but it's more interesting to see example of the specific case.

```{r}
NcGene <- ncRNA_Gene(id = 545, category = "Regolative")
```

### lncRNA

**NB**: The length of the RNA which must be higher than 200. The first example below is going to give an error, just the second is correct.

```{r eval=FALSE}
lncfoxp3 <- lncRNA_Gene(id = 62728, Hugo_symbol = "lncFOXP3", length_lncRNA = 120)
```

```{r}
lncfoxp3 <- lncRNA_Gene(id = 62728, Hugo_symbol = "lncFOXP3", length_lncRNA = 1500)
```

### sncRNA

```{r}
SncGene <- sncRNA_Gene(id = 54865, category = "Regolative", length_sncRNA = 76)
```

#### miRNA

**NB**:

-   the target of the miRNA must be a list
-   for the miRNA the attribute for the length of the product is the general attribute `length_sncRNA` (this is also the same also for `snoRNA_Gene` and `snRNA_Gene`)

```{r}
miRNA1 <- miRNA_Gene(id= 508218, length_sncRNA = 48, RNA_target = list(brca1, lncfoxp3))
```

#### snoRNA

```{r}
SNORD116 <- snoRNA_Gene(id = 8265, description = 'small nucleolar RNA that plays a crucial role in the development of Prader-Willi Syndrome')
```

#### snRNA

```{r}
U1 <- snRNA_Gene(id= 508218, length_sncRNA = 64, gene_product = "snRNA")
```

```{r}
U2 <- snRNA_Gene(id= 57282, length_sncRNA = 68, gene_product = "snRNA")
```

# Functions

## ComputeGeneLength

With this function you can compute the length of the gene in base pair from the start to the end.

```{r}
ComputeGeneLength(tRNA1)
```

```{r}
ComputeGeneLength(rRNA2)
```

## lengthProduct

With this function it's possible to extract the length of the specific product of a gene also without knowing which product is:

```{r}
lengthProduct(lncfoxp3)
```

```{r}
lengthProduct(U1)
```

```{r}
lengthProduct(rRNA4)
```

# Getter and setter
It's easily possible to access attributes of object or modify them:
```{r}
id(brca1)
```

```{r}
RNA_target(miRNA1)
```


# Final recommendation (with wrong examples):

If you will try to run the examples below you will get an error:

-   Don't forget the id:

```{r eval=FALSE}
tp53 <- CodingGene(Hugo_symbol = 'tp53')
```

-   Don't make the location and the chromosome discordant:

```{r eval=FALSE}
MitocondrialGene <- CodingGene(id = 144261, location = 'Nucleus', chrom = 'M')
```

-   Don't make the TSS, TTS and the strand discordant:

```{r eval=FALSE}
smallnoncodingGene <- sncRNA_Gene(id = 9875, strand = '+', TSS = 1023, TTS = 763)
```

-   Don't make the TSS, TTS discordant with the strand and with CDS start and CDS end (just for `CodingeGene`):

```{r eval=FALSE}
pik3ca <- CodingGene(id = 2876, chrom = "chr3", TSS = 980, CDS_start = 900, strand = '+' , number_exon = 21)
```

-   Don't try to change the attributes `category`, `gene_product` and `type_RNA`(only for `ncRNA_Gene`):

```{r eval=FALSE}
snoRNA <- snoRNA_Gene(id = 52719, gene_product = "tRNA")
```

```{r eval=FALSE}
HouseKeepingGene <- Housekeeping_RNA_Gene(id = 6789, category = 'Regolative')
```

```{r eval=FALSE}
smallNuclearRna <- snRNA_Gene(id = 78641, type_RNA = 'long')
```

```{r sessionInfo}
sessionInfo()

```
